A test of diversifying selection for a trait from within and between species variations

### 2023-07-11 manuscript/main.tex
Updated manuscript

\usepackage{url}
\usepackage{amssymb,amsfonts,amsmath,amsthm,mathtools}
\usepackage{lmodern}
\usepackage{xfrac, nicefrac}
\usepackage{bm}
\usepackage{listings, enumerate, enumitem}
\usepackage[export]{adjustbox}
\usepackage{graphicx}
\usepackage{bbold}
\usepackage{pdfpages}
\pdfinclusioncopyfonts=1
\usepackage{lineno}
\usepackage{tabu}
\usepackage{hhline}
\usepackage{multicol,multirow,array}
\usepackage{etoolbox}
\usepackage{booktabs}
\usepackage{makecell}
\usepackage{orcidlink}

\newcommand{\defEqual}{\stackrel{\text{def}}{=}}
\newcommand{\Multiply}{\cdot}
\newcommand{\MultiplyMatrix}{\times}
\newcommand{\UniDimArray}[1]{\bm{#1}}
\newcommand{\BiDimArray}[1]{\bm{#1}}
\newcommand{\tr}{^{\intercal}}
\newcommand{\inv}{^{-1}}
\DeclareMathOperator{\E}{\mathbb{E}}
\DeclareMathOperator{\Var}{\text{Var}}
\DeclareMathOperator{\Cov}{\text{Cov}}
\newcommand{\der}{\mathrm{d}}
\newcommand{\e}{\text{e}}
\newcommand{\Ne}{N_{\text{e}}}
\newcommand{\dn}{d_N}
\newcommand{\ds}{d_S}
\newcommand{\dnds}{\dn / \ds}
\newcommand{\pn}{\pi_N}
\newcommand{\ps}{\pi_S}
\newcommand{\pnps}{\pn / \ps}
\newcommand{\proba}{\mathbb{P}}
\newcommand{\pfix}{\proba_{\text{fix}}}
\newcommand{\Spi}{i}
\newcommand{\Spj}{j}
\newcommand{\NbrTaxa}{n}
\newcommand{\Time}{t}
\newcommand{\Trait}{P}
\newcommand{\Heredity}{h^2}
\newcommand{\HereditySpi}{\Heredity_{\Spi}}
\newcommand{\MeanTrait}{\bar{\Trait_{\Time}}}
\newcommand{\VecTrait}{\UniDimArray{\bar{\Trait}}}
\newcommand{\Root}{0}
\newcommand{\RootTrait}{\widehat{\theta}}
\newcommand{\VarPhy}{\Var \left[\MeanTrait\right]}
\newcommand{\VecOne}{\UniDimArray{1}}
\newcommand{\Distance}{\BiDimArray{D}}
\newcommand{\DistanceMatrix}{\BiDimArray{\Distance}}
\newcommand{\MutationRate}{\mu}
\newcommand{\SubRate}{q}
\newcommand{\NbrLoci}{L}
\newcommand{\VarPhenotype}{V_{\Trait}}
\newcommand{\VarPhenotypeSpi}{V_{\Trait, \Spi}}
\newcommand{\VarGenetic}{V_{\mathrm{G}}}
\newcommand{\VarGeneticSpi}{V_{\mathrm{G}, \Spi}}
\newcommand{\VarEnv}{V_{\mathrm{E}}}
\newcommand{\MatrixGenetic}{\BiDimArray{G}}
\newcommand{\VarMutation}{V_{\mathrm{M}}}
\newcommand{\GenArchi}{\NbrLoci \Multiply \E \left[ a^2 \right]}
\newcommand{\RateMut}{\sigma^2_{\mathrm{M}}}
\newcommand{\RatePhy}{\sigma^2_{\mathrm{B}}}
\newcommand{\RatePop}{\sigma^2_{\mathrm{W}}}
\newcommand{\RatePopSpi}{\sigma^2_{\mathrm{W}, \Spi}}
\newcommand{\VecRatePop}{\UniDimArray{\RatePop}}
\newcommand{\EstRatePhy}{\widehat{\RatePhy}}
\newcommand{\EstRatePop}{\widehat{\RatePop}}
\newcommand{\NI}{\rho}
\newcommand{\StdSelection}{\sigma}
\newcommand{\VarSelection}{\StdSelection^2}

@@

For a given trait, its genetic architecture is defined by the number of loci encoding the trait ($\NbrLoci$) and the random effect of a mutation on the trait ($a$).
New mutations are thus generating variance for such a trait.
At the individual level, the mutational variance ($\VarMutation$) is the rate at which new mutations contribute to the trait variance per generation.
As shown in \textcite{lande_quantitative_1979, lande_sexual_1980}, $\VarMutation$ is a function of the mutation rate per loci per generation ($\MutationRate$), as:
\begin{gather}
    \VarMutation = 2 \MutationRate \Multiply \RateMut \label{eq:var-mutation}.
\end{gather}

@@

Additionally, the genetic variance is equal to the observed phenotypic variance ($\VarPhenotype$) multiplied by heredity ($\Heredity$), giving also $\RatePop$ as:
\begin{align}
    \RatePop & = \frac{\Heredity \Multiply \VarPhenotype }{\pi }. \label{eq:rate-pheno-pop}
\end{align}

@@

@@

Here we generalize to $\Ntrait$ traits evolving along the phylogeny and are correlated between them.
Their variation along the phylogeny is modelled as an $\Ntrait$-dimensional Brownian process $\Brownian$ ($1 \times \Ntrait$) starting a the root and branching along the tree topology.
The rate of change of the Brownian process is determined by the positive semi-definite and symmetric covariance matrix between traits $\CovarianceMatrix$ ($\Ntrait \times \Ntrait$).
We defined the {prior} on the covariance matrix as an inverse Wishart distribution, the {posterior} on $\CovarianceMatrix$, conditional on a particular realization of $\brownian$ is also an invert Wishart distribution.

@@

We will assume that the trait genotypic value is encoded by $\NbrLoci$ loci, with each locus contributing to the genotypic value, these effect are Normally distributed with standard deviation $a$ (Fig.~\ref{fig:simulator}).
We will follow a Wright-Fisher model with mutation, selection and drift to formalise the evolutionary dynamics of the system.
Assuming a diploid panmictic population of size $\Ne$ and with non-overlapping generations, we can simulate explicitly each generation along a phylogenetic tree.
Genetic drift is modelled by the multinomial resampling of the currently segregating alleles.
The number of mutations is drawn from a Poisson distribution with mean $2 \Multiply \MutationRate \Multiply \NbrLoci $.
The changes in the mutation rate per generation ($\MutationRate$) and $\Ne$ along the lineages is modelled by a geometric Brownian process.
Additionally, the instant value of log-$\Ne$ is modelled as a sum of a geometric Brownian process and an Ornstein-Uhlenbeck process.
The geometric Brownian motion accounts for long-term fluctuations, while the Ornstein-Uhlenbeck introduces short-term fluctuations.
The simulations will start from an initial sequence at equilibrium at the root of the tree and, at each node, the process will be split until it finally reaches the leaves of the tree.
From a speciation process perspective, this a equivalent to an allopatric speciation over one generation.
Selection is modelled as a one dimensional Fisher's geometric landscape, with the fitness of an individual being a monotonously decreasing function of the distance between the individual and optimal phenotype \cite{tenaillon_utility_2014,blanquart_epistasis_2016}.
The exact functional phenotype-fitness map depends on $2$ external parameters controlling for strength of selection and optimum phenotype.
Mutations are seen has displacement of the phenotype in the multidimensional space.
Beneficial mutations moves the phenotype closer to the optimum, while deleterious mutations moves it further away.
Stabilizing selection is implemented by means of a fixed optimum phenotype.
Diversifying selection is implemented by means of a moving optimum phenotype, changing along the phylogeny as a geometric Brownian process.

@@

@@

To determine whether a trait is neutrally evolving or under selection, we reviewed the theoretical equations of trait evolution at different scales.
The underlying mutational input for a neutral trait is formally related to both within and between species variation of the trait.
More specifically, for a heritable trait under a neutral regime, the genetic variance of the trait ($\VarGenetic$) normalized by genetic diversity of any neutral genomic region ($\pi$) is a proxy for the mutational variance ($\RatePop$).
At the phylogenetic level, along a specific lineage, the mean value for the trait evolves as a Brownian process, and the variance in this mean value after a time $\Time$ is called $\VarPhy$.
This variance can also be normalized by the genetic distance of any neutral genomic region ($d$), allowing us to obtain another proxy for the mutational variance ($\RatePhy$).
Altogether, we define the neutrality index as $\RatePhy/\RatePop$, which equals to $1$ for a neutral trait, suggesting that traits for which this relationship is not verified are putatively under a selective regime.
Under a depleted variation between species, meaning $\RatePhy < \RatePop$, the trait is putatively under stabilizing, forcing the mean value to be homogeneous between species.
In contrast, under an inflated variation between species, meaning $\RatePhy > \RatePop$, the trait is putatively under diversifying selection, or a moving optimum such that the mean value is heterogeneous in the different species.

\begin{table*}[t!]
    \centering
    \begin{adjustbox}{width = 1.0\textwidth}
    \begin{tabular}{||l|l||c|c||c|c||}
            \hline
            Assumption & Consequences & $\EstRatePop$ & $\EstRatePhy$ & Test $\NI > 1$  & Test $\NI < 1$ \\ \hline \hline
            Changing $\MutationRate$ & Scaling with $\pi$, $d$, $\VarPhenotype$ and $\VarPhy$ & Adjusted & Adjusted & -- & --  \\ \hline
            Changing $\Ne$ & Scaling with $\pi$ and $\VarPhenotype$ & Adjusted & -- & -- & --  \\ \hline
            Unknown heredity & $\Heredity = 1$ is over-estimated  & Over-estimated & -- & Conservative & False positives  \\ \hline
            Inbreeding & $\pi$ is under-estimated  & Over-estimated & -- & Conservative & False positives  \\ \hline
            Polymorphism are selected & $\pi$ is under-estimated  & Over-estimated & -- & Conservative & False positives  \\ \hline
            Sexual dimorphism & $\VarPhenotype$ is over-estimated  & Over-estimated & -- & Conservative & False positives  \\ \hline
            Trait encoded by few loci & $\VarPhy$ is under-estimated  & -- & Under-estimated & Conservative & False positives  \\ \hline
            Divergence are selected & $d$ is under-estimated & -- & Over-estimated & False positives & Conservative  \\ \hline
            Structured population & unknown & unknown & unknown & unknown & unknown \\ \hline
            Gene flow between species & unknown & unknown & unknown & unknown & unknown \\ \hline
    \end{tabular}
    \end{adjustbox}
    \caption{Assumptions and their consequences on deviation from neutraliity.}
    \label{table:assumptions}
\end{table*}

Importantly, the genetic sequences from which $\pi$ and $d$ are estimated should be neutrally evolving, but they are not necessarily linked to the quantitative trait under study, they allow to normalize for diversity driven by mutation rate and population size.
Analogously to the \textcite{mcdonald_adaptative_1991} test devised for protein-coding DNA sequences, a category of neutral mutations (synonymous) is used to determine the neutral expectation, and the inflation of divergence is compared to polymorphism within species.
The relationship between trait variation between and within species can be used as method to detect whether traits are evolving under a neutral model of evolution.
Similarly, in molecular evolution, non-synonymous divergence ($\dn$) and polymorphism $\pn$ are normalized by synonymous divergence ($\ds$) and polymorphism ($\ps$) to produce a neutrality index\cite{mcdonald_adaptative_1991, fay_evaluating_2008}.
Altogether, our neutrality index for a quantitative trait leverages the data for any number of species, and takes advantage of the signal over the whole phylogeny, while at the same time taking into account phylogenetic inertia and addressing the non-independence between species (see fig.~\ref{fig:methods}).
This statistic can be obtained as a maximum likelihood estimate, from eq.~\ref{eq:estimated-rate-pop} and~\ref{eq:estimated-rate-phy}.

@@

\begin{figure*}[!ht]
    \centering
    \includegraphics[width=\textwidth, page=1] {artworks/cst_L5000.rho}
    \includegraphics[width=\textwidth, page=1] {artworks/cst_L5000.pvalues}
    \caption{
        $\RatePhy / \RatePop$ for $1,000$ replicates under different evolutionary regimes.
        Traits are encoded by $5,000$ loci with $a=1$, $h^2=0.2$, $\Ne=50$, $\mu=10^{-5}$, $t=10,000$.
        $\NI < 1$ for traits simulated under selection (stabilizing selection in yellow).
        $\NI = 1$ for traits simulated under a neutral evolution (in blue).
        $\NI > 1$ for genes simulated under a moving optimum (diversifying selection in red).
        Effective population size ($\Ne$) and mutation rate $\MutationRate$ are either constant (top panel), or fluctuating as a Brownian process along the phylogeny (panel B).
    }
    \label{fig:constant_pop_size_phy_pop}
\end{figure*}

@@

In this study, we proposed a neutrality index for a quantitative trait.
At the phylogenetic scale, trait variation between species is normalized by sequence divergence obtained from a neutral set of markers.
Similarly, trait variation within species is normalized by sequence polymorphism obtained from a neutral set of markers.
Our neutrality index for a trait is the ratio of between species variation to within species variation.
Such neutrality index can be compared to 1 to seek a statistical deviation from a neutral trait.
In other words, our statistical test does to seeks to find traits with extremal values in comparison to a pool of traits.
Instead, our procedure can be applied to a single trait, estimating the neutrality index and giving a statistical test for departure from the null model of neutral evolution.
Technically, the neutrality index is estimated either as a maximum likelihood point estimate, or as mean posterior from a Bayesian implementation also giving access to the posterior credible interval allowing to test for a departure from a neutrally evolving trait.
Importantly, our neutrality index leverages sequence divergence and polymorphism, as well as the heritability of traits.
We argue that the main novelty of this study is being able to use sequence divergence and polymorphism to normalize the variation of trait between and within species.
By doing so, our test is not sensitive to the assumption that population sizes ($\Ne$) and mutation rates ($\MutationRate$) are constant across the phylogeny, not requiring to estimate divergence time\cite{litsios_effects_2012}.
Instead, the normalization by sequence divergence and polymorphism automatically absorbs these changes\cite{seo_estimating_2004}.
Contrarily, we argue that the main drawback of our method is that the heritability ($\Heredity$) of a trait is required if one has access only to the phenotypic variance ($\VarPhenotype$) and not the genetic variance ($\VarGenetic$), which is empirically hard to measure for any trait.

>>>>

\usepackage{url}
\usepackage{amssymb,amsfonts,amsmath,amsthm,mathtools}
\usepackage{lmodern}
\usepackage{xfrac, nicefrac}
\usepackage{bm}
\usepackage{listings, enumerate, enumitem}
\usepackage[export]{adjustbox}
\usepackage{graphicx}
\usepackage{bbold}
\usepackage{pdfpages}
\pdfinclusioncopyfonts=1
\usepackage{lineno}
\usepackage{tabu}
\usepackage{hhline}
\usepackage{multicol,multirow,array}
\usepackage{etoolbox}
\usepackage{booktabs}
\usepackage{makecell}
\usepackage{orcidlink}
\usepackage{marvosym}

\newcommand{\defEqual}{\stackrel{\text{def}}{=}}
\newcommand{\Multiply}{\cdot}
\newcommand{\MultiplyMatrix}{\times}
\newcommand{\UniDimArray}[1]{\bm{#1}}
\newcommand{\BiDimArray}[1]{\bm{#1}}
\newcommand{\tr}{^{\intercal}}
\newcommand{\inv}{^{-1}}
\DeclareMathOperator{\E}{\mathbb{E}}
\DeclareMathOperator{\Var}{\text{Var}}
\DeclareMathOperator{\Cov}{\text{Cov}}
\newcommand{\der}{\mathrm{d}}
\newcommand{\e}{\text{e}}
\newcommand{\Ne}{N_{\text{e}}}
\newcommand{\dn}{d_N}
\newcommand{\ds}{d_S}
\newcommand{\dnds}{\dn / \ds}
\newcommand{\pn}{\pi_N}
\newcommand{\ps}{\pi_S}
\newcommand{\pnps}{\pn / \ps}
\newcommand{\proba}{\mathbb{P}}
\newcommand{\pfix}{\proba_{\text{fix}}}
\newcommand{\Spi}{i}
\newcommand{\Spj}{j}
\newcommand{\NbrTaxa}{n}
\newcommand{\Time}{t}
\newcommand{\Trait}{P}
\newcommand{\Heredity}{h^2}
\newcommand{\HereditySpi}{\Heredity_{\Spi}}
\newcommand{\MeanTrait}{\bar{\Trait_{\Time}}}
\newcommand{\VecTrait}{\UniDimArray{\bar{\Trait}}}
\newcommand{\Root}{0}
\newcommand{\RootTrait}{\widehat{\theta}}
\newcommand{\VarPhy}{\Var \left[\MeanTrait\right]}
\newcommand{\VecOne}{\UniDimArray{1}}
\newcommand{\Distance}{\BiDimArray{D}}
\newcommand{\DistanceMatrix}{\BiDimArray{\Distance}}
\newcommand{\MutationRate}{\mu}
\newcommand{\SubRate}{q}
\newcommand{\NbrLoci}{L}
\newcommand{\VarPhenotype}{V_{\Trait}}
\newcommand{\VarPhenotypeSpi}{V_{\Trait, \Spi}}
\newcommand{\VarGenetic}{V_{\mathrm{G}}}
\newcommand{\VarGeneticSpi}{V_{\mathrm{G}, \Spi}}
\newcommand{\VarEnv}{V_{\mathrm{E}}}
\newcommand{\MatrixGenetic}{\BiDimArray{G}}
\newcommand{\VarMutation}{V_{\mathrm{M}}}
\newcommand{\GenArchi}{\NbrLoci \Multiply \E \left[ a^2 \right]}
\newcommand{\RateMut}{\sigma^2_{\mathrm{M}}}
\newcommand{\RatePhy}{\sigma^2_{\mathrm{B}}}
\newcommand{\RatePop}{\sigma^2_{\mathrm{W}}}
\newcommand{\RatePopSpi}{\sigma^2_{\mathrm{W}, \Spi}}
\newcommand{\VecRatePop}{\UniDimArray{\RatePop}}
\newcommand{\EstRatePhy}{\widehat{\RatePhy}}
\newcommand{\EstRatePop}{\widehat{\RatePop}}
\newcommand{\NI}{\rho}
\newcommand{\EstNI}{\widehat{\rho}}
\newcommand{\StdSelection}{\sigma}
\newcommand{\VarSelection}{\StdSelection^2}

@@

For a given trait, its genetic architecture is defined by the number of loci encoding the trait ($\NbrLoci$) and the random effect of a mutation on the trait ($a$).
New mutations are thus generating variance for such a trait.
At the individual level, the mutational variance ($\VarMutation = \NbrLoci \Multiply  \E [a^2]$) is the rate at which new mutations contribute to the trait variance per generation.
As shown in \textcite{lande_quantitative_1979, lande_sexual_1980}, $\VarMutation$ is also a function of the mutation rate per loci per generation ($\MutationRate$), as:
\begin{gather}
    \VarMutation = 2 \MutationRate \Multiply \RateMut \label{eq:var-mutation}.
\end{gather}

@@

Additionally, the genetic variance is equal to the observed phenotypic variance ($\VarPhenotype$) multiplied by heritability ($\Heredity$), giving also $\RatePop$ as:
\begin{align}
    \RatePop & = \frac{\Heredity \Multiply \VarPhenotype }{\pi }. \label{eq:rate-pheno-pop}
\end{align}

@@

Altogether, the neutrality index is estimated as:
\begin{gather}
    \EstNI = \frac{\EstRatePhy}{\EstRatePop}. \label{eq:estimated-NI}
\end{gather}

@@

Even though the neutrality index can be applied for has many traits as available, each trait is so far assumed to be independent.
Assuming covariation between traits is also an extension that can be implemented within the multidimensional Brownian framework~\cite{huelsenbeck_detecting_2003, lartillot_phylogenetic_2011, lartillot_joint_2012, latrille_inferring_2021}.
Here we generalize to $\Ntrait$ traits evolving along the phylogeny and are correlated between them using the \textit{BayesCode} software~\cite{latrille_inferring_2021} .
Their variation along the phylogeny is modelled as an $\Ntrait$-dimensional Brownian process $\Brownian$ ($1 \times \Ntrait$) starting a the root and branching along the tree topology.
The rate of change of the Brownian process is determined by the positive semi-definite and symmetric covariance matrix between traits $\CovarianceMatrix$ ($\Ntrait \times \Ntrait$).
The off-diagonal elements of $\CovarianceMatrix$ are the covariance between traits, and the diagonal elements are the variance of each trait, thus corresponding to $\NI$.
We defined the {prior} on the covariance matrix as an inverse Wishart distribution, the {posterior} on $\CovarianceMatrix$, conditional on a particular realization of $\brownian$ is also an invert Wishart distribution.
The posterior distribution of $\EstNI$ thus allow to test for deviation from neutrality, for example by computing $\proba{\NI > 1 }$ to test for evidence of diversifying selection and $\proba{\NI < 1 }$ to test for evidence of stabilizing selection.

@@

We will assume that the trait genotypic value is encoded by $\NbrLoci$ loci, with each locus contributing to the genotypic value, these effect are Normally distributed with standard deviation $a$ (Fig.~\ref{fig:simulator}).
We will follow a Wright-Fisher model with mutation, selection and drift to formalise the evolutionary dynamics of the system.
Assuming a diploid panmictic population of size $\Ne$ and with non-overlapping generations, we can simulate explicitly each generation along a phylogenetic tree.
Genetic drift is modelled by the multinomial resampling of the currently segregating alleles.
The number of mutations is drawn from a Poisson distribution with mean $2 \Multiply \MutationRate \Multiply \NbrLoci $.
The changes in the mutation rate per generation ($\MutationRate$) and $\Ne$ along the lineages is modelled by a geometric Brownian process.
Additionally, the instant value of log-$\Ne$ is modelled as a sum of a geometric Brownian process and an Ornstein-Uhlenbeck process.
The geometric Brownian motion accounts for long-term fluctuations, while the Ornstein-Uhlenbeck introduces short-term fluctuations.
The simulations will start from an initial sequence at equilibrium at the root of the tree and, at each node, the process will be split until it finally reaches the leaves of the tree.
From a speciation process perspective, this a equivalent to an allopatric speciation over one generation.
Selection is modelled as a one dimensional Fisher's geometric landscape, with the fitness of an individual being a monotonously decreasing function of the distance between the individual and optimal phenotype~\cite{tenaillon_utility_2014,blanquart_epistasis_2016}.
The exact functional phenotype-fitness map depends on $2$ external parameters controlling for strength of selection and optimum phenotype.
Mutations are seen has displacement of the phenotype in the multidimensional space.
Beneficial mutations moves the phenotype closer to the optimum, while deleterious mutations moves it further away.
Stabilizing selection is implemented by means of a fixed optimum phenotype.
Diversifying selection is implemented by means of a moving optimum phenotype, changing along the phylogeny as a geometric Brownian process.
Genetic diversity is measured as the heterozygosity of the neutral markers, and the genetic variance of the trait is measured as the variance of the trait genotypic value.
Genetic divergence is measured as the number of substitutions per site along branches of the phylogeny.
Heritability is estimated from the slope of the regression of offspring trait values on parental trait values\cite{lynch_genetics_1998}, using the information from the last two generations of the simulation.

@@

\subsection{Empirical dataset}

We first focused on a mammalian dataset, with the body size and brain mass as quantitative traits.
For the genetic variation, the mammalian phylogeny is extracted from the Zoonomia project, obtained on a set of neutral markers in \textcite{foley_genomic_2023}, with genetic diversity measured as heterozygosity in \textcite{wilder_contribution_2023}
For quantitative trait variation, the log-transformed body size and brain mass are gathered from \textcite{tsuboi_breakdown_2018}.
We removed individuals not marked as adult, and split the data into males and females.
We discarded species with no intra-specific variability.

We also focused on a dataset of Primates species, with the genetic variation obtained from \textcite{kuderna_global_2023} and the quantitative trait variation also from \textcite{tsuboi_breakdown_2018}, using the same filtering as for the mammalian dataset.
Unfortunately, the primate phylogeny was not obtained on a set of neutral markers as for the mammalian dataset, but across the whole genome.

@@

To determine whether a trait is neutrally evolving or under selection, we reviewed the theoretical equations of trait evolution at different scales.
The underlying mutational input for a neutral trait is formally related to both within and between species variation of the trait.
More specifically, for a heritable trait under a neutral regime, the genetic variance of the trait ($\VarGenetic$) normalized by genetic diversity of any neutral genomic region ($\pi$) is a proxy for the mutational variance ($\RatePop$).
At the phylogenetic level, along a specific lineage, the mean value for the trait evolves as a Brownian process, and the variance in this mean value after a time $\Time$ is called $\VarPhy$.
This variance can also be normalized by the genetic distance of any neutral genomic region ($d$), allowing us to obtain another proxy for the mutational variance ($\RatePhy$).
Altogether, we define the neutrality index as $\NI = \RatePhy/\RatePop$, which equals to $1$ for a neutral trait, suggesting that traits for which this relationship is not verified are putatively under a selective regime.
Under a depleted variation between species, meaning $\NI < 1$, the trait is putatively under stabilizing, forcing the mean value to be homogeneous between species.
In contrast, under an inflated variation between species, meaning $\NI > 1$, the trait is putatively under diversifying selection, or a moving optimum such that the mean value is heterogeneous in the different species.

\begin{table*}[t!]
    \centering
    \begin{adjustbox}{width = 1.0\textwidth}
    \begin{tabular}{||l|l||c|c||c|c||}
            \hline
            Assumption & Consequences & $\EstRatePop$ & $\EstRatePhy$ & Test $\NI > 1$  & Test $\NI < 1$ \\ \hline \hline
            Trait encoded by few loci & $\VarPhy$ is under-estimated  & -- & Under-estimated & Conservative & Invalid  \\ \hline
            Sexual dimorphism & $\VarPhenotype$ is over-estimated  & Over-estimated & -- & Conservative & Invalid  \\ \hline
            Inbreeding & $\pi$ is under-estimated  & Over-estimated & -- & Conservative & Invalid  \\ \hline
            Markers for polymorphism are negatively selected & $\pi$ is under-estimated  & Over-estimated & -- & Conservative & Invalid  \\ \hline
            Markers for divergence are positively selected & $d$ is over-estimated & -- & Under-estimated & Conservative & Invalid  \\ \hline
            Markers for polymorphism are positively selected & $\pi$ is over-estimated  & Under-estimated & -- & Invalid & Conservative  \\ \hline
            Markers for divergence are negatively selected & $d$ is under-estimated & -- & Over-estimated & Invalid & Conservative  \\ \hline
    \end{tabular}
    \end{adjustbox}
    \caption{Assumptions and their consequences on the estimation of $\EstRatePop$ and $\EstRatePhy$ and on the ratio $\NI = \EstRatePhy/\EstRatePop$.
    The two last columns indicate whether the test for diversifying selection ($\NI > 1$) and for stabilizing selection $\NI < 1$ are conservative or invalid due to violated assumptions.
    $\pi$ is the genetic diversity, $d$ is the genetic divergence, $\VarPhenotype$ is the phenotypic variance, $\VarPhy$ is the phylogenetic variance.
    }
    \label{table:assumptions}
\end{table*}

Importantly, the genetic sequences from which $\pi$ and $d$ are estimated should be neutrally evolving, but they are not necessarily linked to the quantitative trait under study, they allow to normalize for diversity driven by mutation rate and population size.
The relationship between trait variation between and within species can be used as method to detect whether traits are evolving under a neutral model of evolution.
Altogether, our neutrality index for a quantitative trait leverages the data for any number of species, and takes advantage of the signal over the whole phylogeny, while at the same time taking into account phylogenetic inertia and addressing the non-independence between species (see fig.~\ref{fig:methods}).
This statistic can be obtained as a maximum likelihood estimate, from eq.~\ref{eq:estimated-rate-pop} and~\ref{eq:estimated-rate-phy}.
We also devised a Bayesian estimated allowing to obtain the posterior distribution of the neutrality index, and to test for evidence of diversifying selection as $\proba [\EstNI > 1]$, and evidence for stabilizing selection as $\proba [\EstNI < 1]$.

@@

\begin{figure*}[!ht]
    \centering
    \begin{minipage}{0.49\textwidth}
        \includegraphics[width=\textwidth, page=1] {artworks/cst_L5000.rho}
    \end{minipage}
    \begin{minipage}{0.49\textwidth}
        \includegraphics[width=\textwidth, page=1] {artworks/cst_L5000.pvalues}
    \end{minipage}
    \begin{minipage}{0.49\textwidth}
        \includegraphics[width=\textwidth, page=1] {artworks/fluNe_L5000.rho}
    \end{minipage}
    \begin{minipage}{0.49\textwidth}
        \includegraphics[width=\textwidth, page=1] {artworks/fluNe_L5000.pvalues}
    \end{minipage}
    \caption{
        $\RatePhy / \RatePop$ for $1,000$ replicates under different evolutionary regimes.
        Traits are encoded by $5,000$ loci with $a=1$, $h^2=0.2$, $\Ne=50$, $\mu=10^{-5}$, $t=10,000$.
        Traits simulated under stabilizing selection in yellow, under a neutral evolution in blue, and under a diver optimum (in red).
        Effective population size ($\Ne$) and mutation rate $\MutationRate$ are either constant (top row), or fluctuating as a Brownian process along the phylogeny (bottom row).
    }
    \label{fig:constant_pop_size_phy_pop}
\end{figure*}

\subsection{Results on empirical data}\label{subsec:results-on-empirical-data}
For mammalian body size, we obtained males (\Male) trait and genetic variation for $\NbrTaxa=36$ species, and in females (\Female) variation for $\NbrTaxa=26$ species.
The posterior estimate of neutrality index are $\EstNI_{\text{\Male}} = 0.346$ ($[0.183,~0.677]$ for $95\%$ credible interval) and $\EstNI_{\text{\Female}} = 0.278$ ($[0.136,~0.639]$ for $95\%$ credible interval).
For mammalian brain mass, we obtained $\EstNI_{\text{\Male}} = 1.375$ ($[0.844,~2.212]$ for $95\%$ credible interval, $\NbrTaxa=36$) and $\EstNI_{\text{\Female}} = 1.762$ ($[1.021,~3.010]$ for $95\%$ credible interval, $\NbrTaxa=26$).
In both males and females, brain mass is under diversifying selection, with p-values of $0.116$ and $0.022$ respectively.

For primates body size, we obtained $\EstNI_{\text{\Male}} = 0.557$ ($[0.406,~0.767]$ for $95\%$ credible interval, $\NbrTaxa=71$) and $\EstNI_{\text{\Female}} = 0.393$ ($[0.277,~0.561]$ for $95\%$ credible interval, $\NbrTaxa=65$).
For primates brain mass, we obtained $\EstNI_{\text{\Male}} = 1.920$ ($[1.397,~2.616]$ for $95\%$ credible interval, $\NbrTaxa=71$) and $\EstNI_{\text{\Female}} = 1.974$ ($[1.393,~2.798]$ for $95\%$ credible interval, $\NbrTaxa=65$).
In both males and females, brain mass is under diversifying selection, with p-values of $0.00$ and $0.00$ respectively.
However, these tests are invalid because the assumption of neutral genetic markers used for normalization is violated (see table~\ref{table:assumptions}).

@@

In this study, we proposed a neutrality index for a quantitative trait.
At the phylogenetic scale, trait variation between species is normalized by sequence divergence obtained from a neutral set of markers.
Similarly, trait variation within species is normalized by sequence polymorphism obtained from a neutral set of markers.
Our neutrality inde for a trait, called $\NI$, is the ratio of between species variation to within species variation.
$\NI$ can be compared to 1 to seek a statistical deviation from a neutral trait.
In other words, our statistical test does to seeks to find traits with extremal values in comparison to a pool of traits.
Instead, our procedure can be applied to a single trait, estimating the neutrality index and giving a statistical test for departure from the null model of neutral evolution.
Technically, the neutrality index is estimated either as a maximum likelihood point estimate, or as mean posterior from a Bayesian implementation also giving access to the posterior credible interval allowing to test for a departure from a neutrally evolving trait (e.g. $ \proba [ \NI > 1 ]$).
Importantly, our neutrality index leverages sequence divergence and polymorphism, as well as the heritability of traits.
We argue that the main novelty of this study is being able to use sequence divergence and polymorphism to normalize the variation of trait between and within species.
By doing so, our test is not sensitive to the assumption that population sizes ($\Ne$) and mutation rates ($\MutationRate$) are constant across the phylogeny, not requiring to estimate divergence time\cite{litsios_effects_2012}.
Instead, the normalization by sequence divergence and polymorphism automatically absorbs these changes\cite{seo_estimating_2004}.
Contrarily, we argue that the main drawback of our method is that the heritability ($\Heredity$) of a trait is required if one has access only to the phenotypic variance ($\VarPhenotype$) and not the genetic variance ($\VarGenetic$), which is empirically hard to measure for any trait.
If heritability is not known, the test for diversifying selection can still be performed (although under-powered) by setting the heritability to 1.0, but the test for stabilizing selection is invalid.
Because our test is based on several assumptions that might not hold on empirical data, we provide a table containing the main assumptions and their consequences on the neutrality index and the test that can be performed (see table~\ref{table:assumptions}).
As example, we can assert that brain size is evolving under diversifying selection in mammals since $\EstNI > 1$ (see fig.~\ref{fig:empirical_results}).
However, at the primate scale the evidence for $\EstNI > 1$ does not necessarily imply that the brain size is evolving under diversifying selection, since the markers used for genetic divergence are not neutral makers which can lead to a spurious $\EstNI > 1$ (see table~\ref{table:assumptions}).

% Assumption that could be relaxed: constant genetic architecture and multidimensional generalization
More theoretically, our development is based on several assumption that could be relaxed.
First, $\RateMut$ is assumed to be constant across the phylogeny, while it is determined by the underlying genetic architecture of a trait, which can be variable among individuals and species.
Having a constant $\RateMut$ across the phylogeny is an assumption that can theoretically by relaxed by adding a time component\cite{arnold_understanding_2008, hohenlohe_mipod_2008}, where $\RateMut(t)$ could jointly be estimated along the phylogeny\cite{kostikova_bridging_2016, gaboriau_multiplatform_2020}.
Also, our Bayesian estimation could integrate uncertainty coming from the estimation of genetic variation, using sequences as input instead of estimated values of polymorphism and divergence.
Altogether, we argue that our neutrality index is a promising statistical procedure allowing to determine the evolutionary regime of a quantitative trait, through articulation of the variance between and within species in a unified framework.
